### 2020-09-10 Manuscript/Response.to.reviewers.tex

\item{Line 178: Spell out harmful algal bloom the first time it?s used.}

{\color{red} RGS: Done }

\item{Line 192: Might be helpful to specifically mention carbon sequestration as an important aspect of sinking rates.}

{\color{red}RGS: Agreed - it will definitely have an effect on the `fate` of the photosynthesized carbon- a quickly sunk carbon will join marine sediments instead of being recycled in teh upper layers of the water masses. We now cite the work of Polimene and colleagues on the role of phytoplankton and carbon sequestration  }

@@

\item{Line 204-205: Citation needed for fish mortality. Eutrophication seems like a cause of blooms, not a result of them. What are the potential cascade effects on benthic-pelagic coupling? The term here on its own isn?t really saying anything.}

{\color{red}RGS: In their work, D. Anderson and colleagues cite a range of consequences of increased HAB blooms, including fish mortality and anoxia - and not eutrophization as we wrongly pointed out. Thanks for noticing. On the second argument - we saw potential consequences in two ways: by affecting recruiting of benthic species with pelagic larvae with reduced diatom diversity/biomass; and by changing the nature of sinking matter for filter feeders. We realize that bentho pelagic coupling is not the only source of carbon for benthic communities in the area of study so (RYAN, check) we have decided to remove this assertion from the discussion }

@@

\item{Line 241: Should be Figure 1C}

{\color{red} Done }

\item{Line 267-268: Which index of abundance from Kelly et al. (2019) was used? That paper presents several and though it declares one to be more useful, I think it would be helpful in this manuscript to state which one was used.}

{\color{red} RGS: We used the eDNAindex (the proportion of reads assigned to each taxa per sample, scaled to the maximum proportion that taxa achieved). We provide the R implementation of this index in the Supplementary material, and have noted so in the text (line 128) }

@@

\item{Line 273-282: It?s hard to discern based on the limited code and data provided exactly how the authors subsampled the different subregions of Puget Sound, since the paper they base their data on has overall values for the whole area. I think this section could use a bit of clarification, and all the code and raw data should be given (the provided github link is broken).}

{\color{red} RGS: We apologize for the broken github code - We have clarified how we simulated the future projection in the Supplementary Information - Also see response to reviewer1 on line 122 - 137 of this letter  }

### 2020-09-10 Manuscript/What.lies.ahead.reviewed.tex
response

We characterized sample carbonate chemistry by measuring Total Alkalinity (TA; open-cell automated titration based on a 876 Dosimat plus (Metrohm AG) as part of a custom system assembled by Andrew Dickson (UCSD) and used in the laboratory of Alex Gagnon at UW) and Dissolved Inorganic Carbon (DIC; Apollo Instruments, USA; CO$_2$ extraction system with 10\% (v/v) phosphoric acid). Both measurements were calibrated and validated with certified reference material from the Scripps Oceanographic Institute. Using DIC and TA, we calculated pH and the remaining carbonate system parameters using the R package `seacarb' \cite{seacarb}.

Our sampled areas differed in the environmental variables driving changes in carbonate chemistry. San Juan Island was less seasonally variable than the Hood Canal in every measured parameter (Figure 1B); the island is more directly affected by summer coastal upwelling as a function of bathymetry and circulation patterns \cite{murray2015inland}, and this appears to be the dominant influence on carbonate chemistry there. By contrast, photosynthesis and respiration likely drive much of the carbonate chemistry variation in the Hood Canal (See Supporting Information).

@@

By treating amplification efficiency as consistent within a given taxon, we created an index of abundance for each taxon across space and time \cite{kelly2019understanding}, using pooled data from technical replicates and mean proportions across biological replicates. We used this index of abundance in the multivariate community analysis, and used binary (presence/absence) data to capture species-level responses to environmental conditions.

@@

 Our results therefore suggest a possible change in relative dominance between diatoms and other phytoplankton species such as dinoflagellates, consistent with those seen at ecological regime shifts found elsewhere \cite{wasmund2017diatom, hinder2012changes}. Such a shift could affect ecosystems in many ways; even under the assumption that the surviving taxa would maintain the primary production levels, for example, the smaller cell-size of dinoflagellates and the differential sinking rates of the two groups would likely alter regional patterns of nutrient cycling \cite{xiao2018warming, anderson2002harmful, bienfang1984sinking}. Although the north Atlantic has shown an increase in diatom abundance \cite{hinder2012changes}, the increase in wind stress and associated mixing in the water column in the open ocean is unlikely to occur in the Hood Canal, where stratification is the strongest in the Salish Sea \cite{moore2008descriptive}. Furthermore, locally focused models support an increase in dinoflagellate dominance with climate change, particularly during summer months \cite{khangaonkar2019salish}.

@@

 Changes in environmental conditions and associated shifts in planktonic communities will likely reshape ecosystems and food webs, although some environmental processes may be conserved even as the particular taxa change. A switch from a diatom-dominated ecosystem to one in which dinoflagellate blooms extend in space and time could provoke cascade effects \cite{xiao2018warming} including fish mortality, eutrophication \cite{anderson2002harmful}, and bentho-pelagic coupling \citep[][]{bienfang1984sinking}. Beyond the phylum-specific patterns, the increase in suitable habitat for harmful algae species will alone be an engine for ecosystem change \cite{trainer2019pelagic, wells2018harmful}.

>>>>

We characterized sample carbonate chemistry by measuring Total Alkalinity (TA; open-cell automated titration based on a 876 Dosimat plus (Metrohm AG) as part of a custom system assembled by Andrew Dickson (UCSD) and used in the laboratory of Alex Gagnon at UW) and Dissolved Inorganic Carbon (DIC; Apollo Instruments, USA; CO$_2$ extraction system with 10\% (v/v) phosphoric acid). Both measurements were calibrated and validated with certified reference material from the Scripps Institution of Oceanography. Using DIC and TA, we calculated pH and the remaining carbonate system parameters using the R package `seacarb' \cite{seacarb}.

Our sampled areas differed in the environmental variables driving changes in carbonate chemistry. San Juan Island was less seasonally variable than the Hood Canal in every measured parameter (Figure 1C); the island is more directly affected by summer coastal upwelling as a function of bathymetry and circulation patterns \cite{murray2015inland}, and this appears to be the dominant influence on carbonate chemistry there. By contrast, photosynthesis and respiration likely drive much of the carbonate chemistry variation in the Hood Canal (See Supporting Information).

@@

By treating amplification efficiency as consistent within a given taxon, we created an index of abundance for each taxon across space and time \cite{kelly2019understanding}, using pooled data from technical replicates and mean proportions across biological replicates. We used this index of abundance in the multivariate community analysis, and used binary (presence/absence) data to capture species-level responses to environmental conditions. {\color{red}An implementation in R of this index is provided in the Supplementary information.}

@@

 Our results therefore suggest a possible change in relative dominance between diatoms and other phytoplankton species such as dinoflagellates, consistent with those seen at ecological regime shifts found elsewhere \cite{wasmund2017diatom, hinder2012changes}. Such a shift could affect ecosystems in many ways; even under the assumption that the surviving taxa would maintain the primary production levels, for example, the smaller cell-size of dinoflagellates and the differential sinking rates of the two groups would likely alter regional patterns of nutrient cycling and carbon sequestration \cite{xiao2018warming, anderson2002harmful, bienfang1984sinking, polimene2017biological}. Although the north Atlantic has shown an increase in diatom abundance \cite{hinder2012changes}, the increase in wind stress and associated mixing in the water column in the open ocean is unlikely to occur in the Hood Canal, where stratification is the strongest in the Salish Sea \cite{moore2008descriptive}. Furthermore, locally focused models support an increase in dinoflagellate dominance with climate change, particularly during summer months \cite{khangaonkar2019salish}.

@@

 Changes in environmental conditions and associated shifts in planktonic communities will likely reshape ecosystems and food webs, although some environmental processes may be conserved even as the particular taxa change. A switch from a diatom-dominated ecosystem to one in which dinoflagellate blooms extend in space and time could provoke cascade effects \cite{xiao2018warming} including fish mortality, anoxia \cite{anderson2002harmful}, and {\color{red} carbon sinking dynamics \st{alter current bentho-pelagic coupling processes} \cite{bienfang1984sinking}}. Beyond the phylum-specific patterns, the increase in suitable habitat for harmful algae species will alone be an engine for ecosystem change \cite{trainer2019pelagic, wells2018harmful}.
